Extended Data

### 2022-03-14 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
Combined with observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
What happens with this extra C?

<!--extra C is probably not going to ANPP_woody-->
Additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$. 
Although $ANPP_{woody}$ is routinely calculated based on stem growth, the two can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited, such that increased GPP does not enhance $ANPP_{woody}$[@delpierre_wood_2016].

<!--on the fate of extra C-->
Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, fail to stimulate $ANPP_{woody}$[@jiang_fate_2020;@refs].
The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, $ANPP_{woody}$ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018].
Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be short lived.

<!--.short vs long-term effects-->
It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005]<!--Can we replace this churkina_2005 ref with another NEE ref? NO, it's making a different point.-->.<!-- The point here is that if you compare across sites in different climates (Harvard Forest and SCBI, for example), you do see greater stem growth associated with the longer growing seasons. Hopefully my rewording makes that clear. -->
Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.

>>>>

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
Combined with widespread observations that warming springs tend to lengthen the season of C sequestration[@lu_no_2022;@finzi_carbon_2020] and increase net C sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion.
What happens with this extra C?

<!--extra C is probably not going to ANPP_woody-->
The presumed additional C fixation in years with warm springs[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020] does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`), and is unlikely to augment $ANPP_{woody}$. 
Although $ANPP_{woody}$ is routinely calculated based on stem growth, the two can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].
It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited[@delpierre_wood_2016;@refs].

<!--on the fate of extra C-->
Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, not cause a corresponding increase in $ANPP_{woody}$[@jiang_fate_2020;@refs].
The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, $ANPP_{woody}$ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018].
Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be short lived.

<!--.short vs long-term effects-->
It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of $ANPP_{woody}$.
Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
If extra photosynthate made available through a growing difference between GPP and $ANPP_{woody}$ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase $ANPP_{woody}$ through indirect mechanisms.
Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.
